---
title: "Housing Prices"
author: "Adrian Luu"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(here)
library(broom)
```

```{r, loading data}
x_test <- read_csv(here::here("data/x_test_houses.csv")) |>
  rename(ID = "(nrow(y_train) + 1):(nrow(y_train) + nrow(y_test))")
y_train <- read_csv(here::here("data/y_train_houses.csv"))
x_train <- read_csv(here::here("data/x_train_houses.csv")) |> 
  rename(ID = "1:nrow(y_train)")

training <- left_join(y_train, x_train, by = "ID") |> select(-ID)
```

## Part (a)

### Linear Regression using basic R-package

`lm()` computes the closed-form OLS solution derived in Lecture 3: minimising $E_{in}(\mathbf{w}) = \frac{1}{N}\|\mathbf{Xw-y}\|^2$ by setting the gradient to zero gives $\mathbf{w}^* = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}$. No iterative algorithm (gradient descent) is needed here since linear regression has a closed-form minimiser, unlike logistic regression or the ANN in part (c).

```{r, impute bedrooms}
bedrooms_median <- median(training$total_bedrooms, na.rm = TRUE)

training <- training |>
  mutate(total_bedrooms = if_else(is.na(total_bedrooms), bedrooms_median, total_bedrooms))

x_test <- x_test |>
  mutate(total_bedrooms = if_else(is.na(total_bedrooms), bedrooms_median, total_bedrooms))
```

```{r, in-sample fit}
final_fit <- lm(Price ~ ., data = training)

rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2)) }

rmse(training$Price, predict(final_fit, training))
summary(final_fit)$r.squared
```

### Linear regression from first principles

`lm()` handles the `ocean_proximity` factor automatically, converting it to dummy columns internally. To build **X** by hand for the pseudo-inverse algorithm above, we have to do that encoding ourselves: one dummy column per category except a baseline (`<1H OCEAN`, which `lm()` also treats as the baseline and absorbs into the intercept).

```{r, manual categorical encoding}
training_encoded <- training |>
  mutate(
    ocean_proximityINLAND      = as.numeric(ocean_proximity == "INLAND"),
    ocean_proximityISLAND      = as.numeric(ocean_proximity == "ISLAND"),
    `ocean_proximityNEAR BAY`   = as.numeric(ocean_proximity == "NEAR BAY"),
    `ocean_proximityNEAR OCEAN` = as.numeric(ocean_proximity == "NEAR OCEAN")
  ) |>
  select(-ocean_proximity)
```

```{r, linear regression algorithm}
X <- cbind(`(Intercept)` = 1, as.matrix(select(training_encoded, -Price)))
y <- training_encoded$Price

w_manual <- solve(crossprod(X), crossprod(X, y))

tibble(
  term = rownames(w_manual),
  manual = as.numeric(w_manual),
  lm = coef(final_fit)[rownames(w_manual)]
)
```

$R^2 = 1 - SS_{res}/SS_{tot}$, where $SS_{res}$ is the model's squared error and $SS_{tot}$ is the squared error from just predicting $\bar{y}$ every time.

```{r, manual r-squared}
predicted_manual <- X %*% w_manual

ss_res <- sum((y - predicted_manual)^2)
ss_tot <- sum((y - mean(y))^2)

r_squared_manual <- 1 - ss_res / ss_tot

r_squared_manual
summary(final_fit)$r.squared
```

```{r, kaggle submission}
test_predictions <- predict(final_fit, newdata = x_test)

submission <- tibble(ID = x_test$ID, Price = test_predictions)

write_csv(submission, here::here("submissions/submission_lm.csv"))
```

| Model             | In-sample R² | Kaggle (out-of-sample) R² |
|-------------------|--------------|---------------------------|
| Linear regression | 0.646        | 0.645                     |

`median_income` has by far the strongest effect: each extra \$10,000 of median block income is associated with roughly \$40,000 higher predicted house price, holding other predictors fixed. `total_rooms` has a *negative* coefficient despite more rooms intuitively meaning a bigger/pricier block - this is a multicollinearity artefact, not a genuine effect: `total_rooms`, `total_bedrooms`, `population`, and `households` are all block-level totals that move together, so the model can't cleanly separate their individual contributions. This is the kind of issue the ratio features in part (b) are meant to fix.
