---
title: "Housing Prices"
author: "Adrian Luu"
format: html
editor: visual
---

```{r}
library(tidyverse)
library(here)
library(broom)
```

```{r, loading data}
x_test <- read_csv(here::here("data/x_test_houses.csv")) |>
  rename(ID = "(nrow(y_train) + 1):(nrow(y_train) + nrow(y_test))")
y_train <- read_csv(here::here("data/y_train_houses.csv"))
x_train <- read_csv(here::here("data/x_train_houses.csv")) |>
  rename(ID = "1:nrow(y_train)")

training <- left_join(y_train, x_train, by = "ID") |> select(-ID)
```

## EDA

Before engineering any features, look at the raw data to see what patterns and issues are actually present.

### Distribution of ocean_proximity

```{r, eda ocean proximity counts}
training |>
  count(ocean_proximity) |>
  ggplot(aes(x = reorder(ocean_proximity, n), y = n)) +
  geom_col() +
  coord_flip() +
  labs(title = "Distribution of ocean_proximity", x = NULL, y = "Count")
```

- Category sizes are wildly unequal: `<1H OCEAN` (6,835 rows) and `INLAND` (4,903) dominate, while `ISLAND` has only 4 rows in the entire training set
- Matters for how much to trust any group-level comparison involving `ISLAND` later on

### Distributions of all continuous variables

```{r, eda all distributions}
training |>
  select(longitude, latitude, housing_median_age, total_rooms, total_bedrooms,
         population, households, median_income, Price) |>
  pivot_longer(everything(), names_to = "variable", values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_histogram(bins = 40) +
  facet_wrap(~variable, scales = "free", ncol = 3) +
  labs(title = "Distributions of all continuous variables", x = NULL, y = "Count")
```

- `Price`: right-skewed with a large spike at exactly \$500,001 (723 of 15,480 rows) - a top-coded/capped value, not a genuine cluster of identical home prices. Motivates modelling `log(Price)` in part (b), and explains the diagonal-line artifact that shows up in residual plots regardless of model, since a fixed number of rows share an identical true value while their predictions vary
- `housing_median_age`: a spike at exactly 52, the maximum (956 rows) - top-coded ("52+ years"). Motivates the `age_capped` flag in part (b)
- `median_income`: also top-coded, at exactly 15.0001 (33 rows) - the same pattern as `Price` and `housing_median_age`, just affecting far fewer rows (0.2%, vs. 4.7% and 6.2%), so not worth a dedicated feature the way `age_capped` was
- `total_rooms`, `total_bedrooms`, `population`, `households`: all heavily right-skewed with a sharp peak near small values and a long tail - motivates the `log_total_rooms` etc. features in part (b)
- `longitude`, `latitude`: multi-modal, with distinct peaks around the SF Bay Area and LA - reflects population being concentrated in a few metro areas rather than spread evenly

### Raw totals vs. households per block

```{r, eda totals vs households}
training |>
  select(households, total_rooms, total_bedrooms, population) |>
  pivot_longer(-households, names_to = "variable", values_to = "value") |>
  ggplot(aes(x = households, y = value)) +
  geom_point(alpha = 0.15, size = 0.6) +
  facet_wrap(~variable, scales = "free_y") +
  labs(title = "Raw totals vs. households per block", x = "Households", y = "Total")
```

- `total_rooms`, `total_bedrooms`, and `population` all sit almost exactly on a straight line through the origin against `households` - these "totals" are overwhelmingly just measuring how many households are in the block, not anything about the housing itself
- A block's raw total is mechanically proportional to its size, so the genuine signal about housing characteristics is in the *ratio* to `households`, not the raw total - this is a stronger justification for the bespoke ratio features than correlation between the totals alone would be

### Price by location

```{r, eda price location}
ca_map <- map_data("state") |> filter(region == "california")

training |>
  ggplot() +
  geom_polygon(data = ca_map, aes(x = long, y = lat, group = group), fill = "grey95", colour = "grey50") +
  geom_point(aes(x = longitude, y = latitude, colour = Price), alpha = 0.4, size = 0.8) +
  scale_colour_viridis_c() +
  coord_quickmap() +
  labs(title = "Price by location", x = "Longitude", y = "Latitude")
```

- Price clearly varies by location, following California's coastline rather than either axis alone
- Motivates the polynomial/interaction location terms in part (b) (`longitude^2`, `latitude^2`, `longitude*latitude`), since a plain linear term in `longitude`/`latitude` alone can only fit an axis-aligned gradient

### Mean Price by ocean_proximity

```{r, eda price by ocean proximity}
training |>
  group_by(ocean_proximity) |>
  summarise(
    n = n(),
    mean_price = mean(Price),
    se_price = sd(Price) / sqrt(n)
  ) |>
  mutate(lower = mean_price - 1.96 * se_price, upper = mean_price + 1.96 * se_price) |>
  ggplot(aes(x = reorder(ocean_proximity, mean_price), y = mean_price)) +
  geom_pointrange(aes(ymin = lower, ymax = upper)) +
  labs(title = "Mean Price by ocean_proximity (95% CI)", x = "ocean_proximity", y = "Mean Price ($)")
```

- Mean price, not a boxplot of the raw block values, because `Price` is already a block-level aggregate and category sizes are wildly unequal (see above) - the CI width directly reflects how much each group's estimate should be trusted
- `INLAND` sits clearly lowest with a tight CI (n = 4,903); the three ocean-adjacent categories cluster around \$240-258k; `ISLAND` has the highest mean (\$403,675) but a CI spanning \$334k-\$473k, since it's estimated from only 4 rows
- Motivates keeping `ocean_proximity` as a predictor, and treating `ISLAND`'s estimated effect with caution given how little data supports it

### Mean Price by median_income decile and ocean_proximity

```{r, eda price by income and ocean proximity}
training |>
  mutate(income_decile = ntile(median_income, 10)) |>
  group_by(ocean_proximity, income_decile) |>
  summarise(n = n(), mean_price = mean(Price), se_price = sd(Price) / sqrt(n), .groups = "drop") |>
  mutate(lower = mean_price - 1.96 * se_price, upper = mean_price + 1.96 * se_price) |>
  filter(n >= 10) |>
  ggplot(aes(x = income_decile, y = mean_price, colour = ocean_proximity, fill = ocean_proximity)) +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.15, colour = NA) +
  geom_line() +
  geom_point(size = 1.5) +
  labs(title = "Mean Price by median_income decile and ocean_proximity",
       x = "median_income decile", y = "Mean Price ($)", colour = "ocean_proximity", fill = "ocean_proximity")
```

- `ISLAND` drops out automatically (only 4 rows total, too few to support even one income decile with `n >= 10`) - consistent with the wide CI above
- `INLAND` sits below every coastal category at every income level, and the gap *widens* with income rather than staying constant - the income-price slope is less steep for `INLAND` than for the coastal categories
- Direct evidence for the `median_income:ocean_proximity` interaction term in part (b): this isn't just "location matters," it's "location changes how much income matters"

## Part (a)

### Linear Regression using basic R-package

`lm()` computes the closed-form OLS solution derived in Lecture 3: minimising $E_{in}(\mathbf{w}) = \frac{1}{N}\|\mathbf{Xw-y}\|^2$ by setting the gradient to zero gives $\mathbf{w}^* = (\mathbf{X}^T\mathbf{X})^{-1}\mathbf{X}^T\mathbf{y}$. No iterative algorithm (gradient descent) is needed here since linear regression has a closed-form minimiser, unlike logistic regression or the ANN in part (c).

```{r, impute bedrooms}
bedrooms_median <- median(training$total_bedrooms, na.rm = TRUE)

training <- training |>
  mutate(total_bedrooms = if_else(is.na(total_bedrooms), bedrooms_median, total_bedrooms))

x_test <- x_test |>
  mutate(total_bedrooms = if_else(is.na(total_bedrooms), bedrooms_median, total_bedrooms))
```

```{r, in-sample fit}
final_fit <- lm(Price ~ ., data = training)

rmse <- function(actual, predicted) {
  sqrt(mean((actual - predicted)^2)) }

rmse(training$Price, predict(final_fit, training))
summary(final_fit)$r.squared
```

```{r, resid vs fitted part a}
resid_plot <- function(fitted, actual, title) {
  tibble(fitted = fitted, resid = actual - fitted) |>
    ggplot(aes(x = fitted, y = resid)) +
    geom_point(alpha = 0.15, size = 0.5) +
    geom_hline(yintercept = 0, colour = "red", linetype = "dashed") +
    labs(title = title, x = "Fitted", y = "Residual")
}

resid_plot(predict(final_fit, training), training$Price, "Residuals vs fitted: Linear regression (part a)")
```

### Linear regression from first principles

`lm()` handles the `ocean_proximity` factor automatically, converting it to dummy columns internally. To build **X** by hand for the pseudo-inverse algorithm above, we have to do that encoding ourselves: one dummy column per category except a baseline (`<1H OCEAN`, which `lm()` also treats as the baseline and absorbs into the intercept).

```{r, manual categorical encoding}
training_encoded <- training |>
  mutate(
    ocean_proximityINLAND      = as.numeric(ocean_proximity == "INLAND"),
    ocean_proximityISLAND      = as.numeric(ocean_proximity == "ISLAND"),
    `ocean_proximityNEAR BAY`   = as.numeric(ocean_proximity == "NEAR BAY"),
    `ocean_proximityNEAR OCEAN` = as.numeric(ocean_proximity == "NEAR OCEAN")
  ) |>
  select(-ocean_proximity)
```

```{r, linear regression algorithm}
X <- cbind(`(Intercept)` = 1, as.matrix(select(training_encoded, -Price)))
y <- training_encoded$Price

w_manual <- solve(crossprod(X), crossprod(X, y))

tibble(
  term = rownames(w_manual),
  manual = as.numeric(w_manual),
  lm = coef(final_fit)[rownames(w_manual)]
)
```

$R^2 = 1 - SS_{res}/SS_{tot}$, where $SS_{res}$ is the model's squared error and $SS_{tot}$ is the squared error from just predicting $\bar{y}$ every time.

```{r, manual r-squared}
predicted_manual <- X %*% w_manual

ss_res <- sum((y - predicted_manual)^2)
ss_tot <- sum((y - mean(y))^2)

r_squared_manual <- 1 - ss_res / ss_tot

r_squared_manual
summary(final_fit)$r.squared
```

```{r, kaggle submission}
test_predictions <- predict(final_fit, newdata = x_test)

submission <- tibble(ID = x_test$ID, Price = test_predictions)

write_csv(submission, here::here("submissions/submission_lm.csv"))
```

| Model             | In-sample R² | Kaggle (out-of-sample) R² |
|-------------------|--------------|---------------------------|
| Linear regression | 0.646        | 0.645                     |

`median_income` has by far the strongest effect: each extra \$10,000 of median block income is associated with roughly \$40,000 higher predicted house price, holding other predictors fixed. `total_rooms` has a *negative* coefficient despite more rooms intuitively meaning a bigger/pricier block - this is a multicollinearity artefact, not a genuine effect: `total_rooms`, `total_bedrooms`, `population`, and `households` are all block-level totals that move together, so the model can't cleanly separate their individual contributions. This is the kind of issue the ratio features in part (b) are meant to fix.

## Part (b)

Part (b) builds four models on top of part (a), each building on the last: (1) engineered features fit with plain OLS, (2) the same features with ridge regularisation to address the collinearity the features introduce, (3) a log(Price) target transform to address the right-skew in `Price` itself, and (4) ridge regularisation combined with the log(Price) target, to check whether the two techniques help more together than separately. Model 3 is the final part (b) model.

### Feature engineering: bespoke and polynomial features

- `rooms_per_household`, `bedrooms_per_room`, `population_per_household`, `rooms_per_person` (Lecture 3 bespoke ratios) - directly target the multicollinearity from part (a), where `total_rooms`, `total_bedrooms`, `population`, and `households` were fighting each other as raw block-level totals
- `median_income_sq`, `housing_median_age_sq`, `income_x_age`, `longitude_x_latitude`, `longitude_sq`, `latitude_sq` (Lecture 5 polynomial/interaction terms) - complete the canonical quadratic basis $z = (1, x_1, x_2, x_1x_2, x_1^2, x_2^2)$ for `median_income`/`housing_median_age` and for `longitude`/`latitude`; the spatial terms let the model fit a diagonal price gradient along California's coastline, which `longitude`/`latitude` alone (axis-aligned only) cannot
- `log_total_rooms`, `log_total_bedrooms`, `log_population`, `log_households` - correct the heavy right-skew in those raw counts, which was producing extreme leverage points and negative predicted prices in part (a)'s residuals
- `dist_to_city` - Euclidean distance (in degrees) from each block to the nearest of SF/LA/San Diego/San Jose, a geometric proxy for distance to a major job/price centre
- `age_capped` - flags the 956 of 15,480 training rows where `housing_median_age` is top-coded at exactly 52 ("52+ years" rather than a precise age)
- `median_income:ocean_proximity` (added in the model formula, not as a precomputed column) - lets income's effect on price vary by location type
- `bedrooms_per_room` is capped at 1: bedrooms can't outnumber total rooms, and real (non-imputed) data never breaches that ratio, but a few rows have median-imputed `total_bedrooms` sitting on an atypically small block, which can push the *uncapped* ratio above 1 and produce unrealistic predictions further down the pipeline

```{r, bespoke and polynomial features}
city_coords <- tribble(
  ~city,       ~city_lon,  ~city_lat,
  "SF",        -122.4194,  37.7749,
  "LA",        -118.2437,  34.0522,
  "San Diego", -117.1611,  32.7157,
  "San Jose",  -121.8863,  37.3382
)

dist_to_nearest_city <- function(longitude, latitude) {
  dists <- sapply(seq_len(nrow(city_coords)), function(i) {
    sqrt((longitude - city_coords$city_lon[i])^2 + (latitude - city_coords$city_lat[i])^2)
  })
  apply(dists, 1, min)
}

engineer_features <- function(df) {
  df |>
    mutate(
      rooms_per_household = total_rooms / households,
      bedrooms_per_room = pmin(total_bedrooms / total_rooms, 1),
      population_per_household = population / households,
      rooms_per_person = total_rooms / population,
      median_income_sq = median_income^2,
      housing_median_age_sq = housing_median_age^2,
      income_x_age = median_income * housing_median_age,
      longitude_x_latitude = longitude * latitude,
      longitude_sq = longitude^2,
      latitude_sq = latitude^2,
      log_total_rooms = log(total_rooms),
      log_total_bedrooms = log(total_bedrooms),
      log_population = log(population),
      log_households = log(households),
      age_capped = as.numeric(housing_median_age == 52),
      dist_to_city = dist_to_nearest_city(longitude, latitude)
    ) |>
    select(-total_rooms, -total_bedrooms, -population, -households)
}

training <- engineer_features(training)
x_test <- engineer_features(x_test)
```

### Standardisation

Per Lecture 5, features on very different scales (e.g. `total_rooms` in the thousands vs `median_income` in single digits) get arbitrarily different weight magnitudes, which matters once regularisation penalises `||w||`. Standardise every continuous predictor to mean 0, variance 1 using *training* statistics only, applied to both training and test.

```{r, standardise}
continuous_vars <- c(
  "longitude", "latitude", "housing_median_age", "median_income",
  "rooms_per_household", "bedrooms_per_room", "population_per_household",
  "rooms_per_person", "median_income_sq", "housing_median_age_sq", "income_x_age",
  "longitude_x_latitude", "longitude_sq", "latitude_sq",
  "log_total_rooms", "log_total_bedrooms", "log_population", "log_households",
  "dist_to_city"
)

train_means <- colMeans(training[continuous_vars])
train_sds <- sapply(training[continuous_vars], sd)

for (col in continuous_vars) {
  training[[col]] <- (training[[col]] - train_means[col]) / train_sds[col]
  x_test[[col]] <- (x_test[[col]] - train_means[col]) / train_sds[col]
}
```

### Model 1: linear regression on engineered features

```{r, refit part b}
fit_b <- lm(Price ~ . + median_income:ocean_proximity, data = training)

rmse(training$Price, predict(fit_b, training))
summary(fit_b)$r.squared
```

```{r, kaggle submission part b}
test_predictions_b <- predict(fit_b, newdata = x_test)

submission_b <- tibble(ID = x_test$ID, Price = test_predictions_b)

write_csv(submission_b, here::here("submissions/submission_b.csv"))
```

```{r, resid vs fitted model 1}
resid_plot(predict(fit_b, training), training$Price, "Residuals vs fitted: Model 1 (engineered features, OLS)")
```

With `longitude`, `latitude`, `longitude^2`, `latitude^2`, and `longitude*latitude` all in the model together, the individual coefficients on these terms are large and unstable - they're highly collinear by construction, all five being functions of the same two raw variables over a narrow real-world range - even though the overall fit and predictions are fine. This is the exact symptom Lecture 5's ridge penalty is designed to fix, which motivates Model 2.

### Model 2: ridge regression with cross-validated lambda

Lecture 5's augmented error adds a ridge penalty to the loss: $E_{aug}(\mathbf{w}) = E_{in}(\mathbf{w}) + \frac{\lambda}{N}\|\mathbf{w}\|_2^2$, which also has a closed-form solution, $\mathbf{w}^*_{\text{ridge}} = (\mathbf{X}^T\mathbf{X} + \lambda I)^{-1}\mathbf{X}^T\mathbf{y}$ - the same normal equations as part (a) with $\lambda I$ added (the intercept is excluded from the penalty, standard practice, since there's no reason to shrink the baseline level toward zero). $\lambda$ is chosen by k-fold cross-validation, per the "Setting $\lambda$" / "Cross-Validation" slides: for each candidate $\lambda$ on a grid, fit on $K-1$ folds and evaluate on the held-out fold, then pick the $\lambda$ with the smallest average out-of-sample error.

```{r, ridge regression function}
X <- model.matrix(Price ~ . + median_income:ocean_proximity, data = training)
y <- training$Price
X_test <- model.matrix(~ . + median_income:ocean_proximity, data = select(x_test, -ID))

ridge_fit <- function(X, y, lambda) {
  penalty <- lambda * diag(ncol(X))
  penalty[1, 1] <- 0  # don't penalise the intercept
  solve(t(X) %*% X + penalty, t(X) %*% y)
}
```

```{r, cross-validate lambda}
set.seed(3555)
k <- 5
folds <- sample(rep(1:k, length.out = nrow(X)))
lambda_grid <- 10^seq(-4, 4, length.out = 20)
cv_rmse <- numeric(length(lambda_grid))

for (i in seq_along(lambda_grid)) {
  lambda <- lambda_grid[i]
  fold_rmse <- numeric(k)

  for (fold in 1:k) {
    is_train <- folds != fold
    is_valid <- folds == fold

    w <- ridge_fit(X[is_train, ], y[is_train], lambda)
    predicted <- X[is_valid, ] %*% w

    fold_rmse[fold] <- rmse(y[is_valid], predicted)
  }

  cv_rmse[i] <- mean(fold_rmse)
}

best_lambda <- lambda_grid[which.min(cv_rmse)]
best_lambda

tibble(lambda = lambda_grid, cv_rmse = cv_rmse) |>
  ggplot(aes(x = lambda, y = cv_rmse)) +
  geom_line() +
  geom_point() +
  scale_x_log10() +
  geom_vline(xintercept = best_lambda, linetype = "dashed", colour = "red") +
  theme_minimal() +
  labs(x = "lambda (log scale)", y = "5-fold CV RMSE", title = "Cross-validated ridge lambda")
```

```{r, ridge fit and kaggle submission}
w_ridge <- ridge_fit(X, y, best_lambda)

rmse(y, X %*% w_ridge)
1 - sum((y - X %*% w_ridge)^2) / sum((y - mean(y))^2)

test_predictions_ridge <- as.numeric(X_test %*% w_ridge)

submission_ridge <- tibble(ID = x_test$ID, Price = test_predictions_ridge)

write_csv(submission_ridge, here::here("submissions/submission_b_ridge.csv"))
```

```{r, resid vs fitted model 2}
resid_plot(as.numeric(X %*% w_ridge), training$Price, "Residuals vs fitted: Model 2 (ridge)")
```

**Findings:**

- CV selects $\lambda \approx 10^{-4}$, i.e. essentially no regularisation ($\lambda \to 0 \approx$ OLS).
- This doesn't contradict the coefficient instability noted above - the individual coefficients on the five collinear spatial terms are still large and poorly determined.
- It means that with $n = 15{,}480$ training rows and only $p = 29$ parameters, there isn't enough variance in the *predictions* (as opposed to the individual coefficients) for shrinkage to pay off - in-sample R² for this model comes out effectively identical to Model 1 (0.694 vs 0.694).
- Kaggle score for this model (`submission_b_ridge.csv`): 0.690 R².
- Ridge is still the right tool to reach for as feature/parameter count grows (particularly once part (c)'s ANN is in play) - it just isn't doing much work for this specific linear model, so Model 3 below builds on Model 1 (plain OLS) rather than on the ridge fit. The same `ridge_fit()` function, lambda grid, and fold assignment are reused again in Model 4, refit on the log(Price) target, to check whether combining the two techniques changes anything.

### Model 3: log-transformed target

House prices are right-skewed, and part (a)'s residual plot showed a funnel shape (errors growing with predicted price) - both point to modelling `log(Price)` instead of `Price` directly. This is the final part (b) model.

```{r, log price model}
fit_log <- lm(log(Price) ~ . + median_income:ocean_proximity, data = training)

predicted_log <- predict(fit_log, training)

log_resid <- log(training$Price) - predicted_log
smear <- mean(exp(log_resid))

predicted_price <- exp(predicted_log) * smear

rmse(training$Price, predicted_price)
1 - sum((training$Price - predicted_price)^2) / sum((training$Price - mean(training$Price))^2)
```

```{r, kaggle submission log price}
test_pred_log <- predict(fit_log, newdata = x_test)
test_pred_price <- exp(test_pred_log) * smear

submission_log <- tibble(ID = x_test$ID, Price = test_pred_price)

write_csv(submission_log, here::here("submissions/submission_b_log.csv"))
```

Since the model is fit on `log(Price)`, the residual diagnostic belongs on that same scale, not the back-transformed dollar scale.

```{r, resid vs fitted model 3}
resid_plot(predicted_log, log(training$Price), "Residuals vs fitted (log scale): Model 3 (log target, final)")
```

**Notes:**

- Naive back-transform (`exp(predicted_log)`) is biased low - by Jensen's inequality, $E[\exp(X)] \geq \exp(E[X])$, so exponentiating a prediction of the *mean* of `log(Price)` underestimates the mean of `Price`. `smear` (Duan's smearing estimator: the average of `exp(residual)` on the log scale) corrects for this.
- Every prediction is `exp(...)`, so they're guaranteed positive - this resolves the negative-predicted-price extrapolation issue flagged in part (a).
- Builds on Model 1's feature set (not Model 2's ridge fit), since Model 2 found regularisation unnecessary here.

### Model 4: ridge regression on the log(Price) target

As a further check, Model 2's ridge machinery (the same `ridge_fit()` function, lambda grid, and fold assignment) is reused here, refit on the log(Price) target rather than raw `Price`, to see whether combining regularisation with the log transform beats either technique alone.

```{r, ridge log price cv}
y_log <- log(training$Price)
cv_rmse_log <- numeric(length(lambda_grid))

for (i in seq_along(lambda_grid)) {
  lambda <- lambda_grid[i]
  fold_rmse <- numeric(k)

  for (fold in 1:k) {
    is_train <- folds != fold
    is_valid <- folds == fold

    w <- ridge_fit(X[is_train, ], y_log[is_train], lambda)
    predicted <- X[is_valid, ] %*% w

    fold_rmse[fold] <- rmse(y_log[is_valid], predicted)
  }

  cv_rmse_log[i] <- mean(fold_rmse)
}

best_lambda_log <- lambda_grid[which.min(cv_rmse_log)]
best_lambda_log
```

```{r, ridge log price model}
w_ridge_log <- ridge_fit(X, y_log, best_lambda_log)
predicted_log_ridge <- X %*% w_ridge_log

log_resid_ridge <- y_log - predicted_log_ridge
smear_ridge <- mean(exp(log_resid_ridge))

predicted_price_ridge_log <- exp(predicted_log_ridge) * smear_ridge

rmse(training$Price, predicted_price_ridge_log)
1 - sum((training$Price - predicted_price_ridge_log)^2) / sum((training$Price - mean(training$Price))^2)
```

```{r, kaggle submission ridge log price}
test_pred_log_ridge <- X_test %*% w_ridge_log
test_pred_price_ridge_log <- as.numeric(exp(test_pred_log_ridge) * smear_ridge)

submission_ridge_log <- tibble(ID = x_test$ID, Price = test_pred_price_ridge_log)

write_csv(submission_ridge_log, here::here("submissions/submission_b_ridge_log.csv"))
```

```{r, resid vs fitted model 4}
resid_plot(as.numeric(predicted_log_ridge), y_log, "Residuals vs fitted (log scale): Model 4 (ridge + log target)")
```

**Findings:**

- CV on the log scale again selects $\lambda \approx 10^{-4}$ ($\approx 0$) - regularisation doesn't help here either.
- In-sample fit is virtually identical to Model 3's plain OLS fit on log(Price) (RMSE 62,212 either way, R² 0.709) - confirms Model 2's finding holds regardless of the target's scale.
- Combining the two techniques doesn't outperform either alone here, so Model 3 (simpler, no regularisation needed) stays the designated final model - confirmed out-of-sample too: Model 4's Kaggle score (`submission_b_ridge_log.csv`) is 0.688, identical to Model 3's.

| Model | In-sample R² | Kaggle (out-of-sample) R² |
|------------------------|------------------------|------------------------|
| Linear regression (part a) | 0.646 | 0.645 |
| Model 1: engineered features (OLS) | 0.694 | 0.690 |
| Model 2: + ridge regression, CV-selected lambda | 0.694 | 0.690 |
| Model 3: + log(Price) target, smeared back-transform (final) | 0.709 (RMSE 62,212) | 0.688 |
| Model 4: ridge regression on log(Price) target | 0.709 (RMSE 62,212) | 0.688 |
